Think deep for reef resilience

### 2025-08-15 paper/paper_draft.qmd
Add important information on coral groups that are represented
Added to supplementary incase people are looking to see what groups we use data from.
Fix typos
variations of "maintainence" to maintenance

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef cover timeseries and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintainence of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-timeseries resilience and
    maintainence of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Our findings suggest that depth provides critical
    protection against thermal stress, contrasting current restoration practices that primarily
    target easily accessible shallow sites. These results have important implications for
    reef management, indicating that depth refugia should be prioritized for further
    consideration in conservation planning alongside connectivity considerations to maximize
    coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
editor:
  markdown:
    wrap: 92
---

@@

Reefs with greater median depth levels maintain coral cover above indicative carbonate
budget levels for a longer time frame compared to reefs with shallower median depths
[@fig-access-esm1-5-depth-carbonate]. Reef median depth has a moderate positive correlation
with the number of years a reef maintains coral cover above the required threshold
[@fig-access-esm1-5-depth-carbonate]. Depth - carbonate budget maintainence correlation
decreases as $\theta_{0}$ increases. This is likely due to an increasing number of reefs
being unable to maintain coral cover above threshold levels as the threshold increases, with
values becoming increasingly right-skewed. The base-10 log of total connectivity strength
was found to have no correlation with the number of years a reef maintains cover above the
required thresholds [@sfig-access-esm1-5-conn-carbonate]. Log connectivity strength -
carbonate budget maintainence correlation slightly decreases as the required carbonate
budget threshold increases.

@@

NorESM2-MM has the lowest Equilibrium Climate Sensitivity (2.5°C) and reefs exhibit
continuing recovery and decline dynamics up to 2075 under this possible climate future.
Recovery phases are predominantly seen in the ‘high’ cover cluster. The ‘medium’ cluster,
and to a lesser degree the ‘low’ cluster, also exhibit periods of recovery and increasing
proportions of coral cover up to 2045. Under NorESM2-MM the Far Northern management area has
lower CID distances between low and high cover clusters than in other management areas
[@tbl-cluster-cid], this is reflected in the low degree of cover maintainence
[@fig-gcm-cover-trajectories]. ACCESS-ESM1-5 has an intermediate ECS value of 3.87°C which
results in intermediate levels of coral cover maintenance (between those of lower and higher
ECS GCMs) [@fig-gcm-cover-trajectories], reflected in intermediate CID distance values
within most management areas [@tbl-cluster-cid]. In Townsville-Whitsunday and Far Northern
management areas, reefs in the high cluster are able to maintain coral cover levels around
20% to 2050 under ACCESS-ESM1-5. Under ACCESS-CM2 (4.66°C), EC-Earth3-Veg (4.30°C) and
GFDL-CM4 (4.1°C) GCMs cover trajectories are largely declining in all management areas,
reflected in lower CID distance values than other GCMs [@tbl-cluster-cid].

@@

Under all GCMs many reefs are projected to suffer substantial declines in coral cover
relative to 2025 levels, with an average of 38% of reefs (924 reefs) clustered into the
lowest cover level cluster at the management area scale across GCMs. NorESM2-MM and
ACCESS-ESM1-5 GCMs convey the most coral cover maintenance into the future. Differences
between low, medium and high reef clusters in cover timeseries are most notable under these
two GCMs [@tbl-cluster-cid; @fig-gcm-cover-trajectories]. Maintainance of coral cover into
the future is linked to the Equilibrium Climate Sensitivity of the GCM used as levels of
future recovery and timespan of cover maintainance increase as GCM ECS increases
[@fig-gcm-cover-trajectories]. Under high-ECS GCMs (GFDL-CM4, EC-Earth3-Veg and ACCESS-CM2)
large declines and lower rates of recovery in high cover clusters are common compared to
lower-ECS GCMs.

@@

CoralBlox uses user-defined size classes that influence size-dependent growth and mortality
rates, through which the coral colonies progress as they grow in diameter. Within ADRIA, we
define the distributions of these parameters, and the size classes based on ecological
monitoring data from the Ecological Intelligence for Reef Restoration Program (EcoRRAP)
([@australianinstituteofmarinescience2025; @ribeirodealmeida2024]).

>>>>

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef cover timeseries and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-timeseries resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Our findings suggest that depth provides critical
    protection against thermal stress, contrasting current restoration practices that primarily
    target easily accessible shallow sites. These results have important implications for
    reef management, indicating that depth refugia should be prioritized for further
    consideration in conservation planning alongside connectivity considerations to maximize
    coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
editor:
  markdown:
    wrap: 92
---

@@

Reefs with greater median depth levels maintain coral cover above indicative carbonate
budget levels for a longer time frame compared to reefs with shallower median depths
[@fig-access-esm1-5-depth-carbonate]. Reef median depth has a moderate positive correlation
with the number of years a reef maintains coral cover above the required threshold
[@fig-access-esm1-5-depth-carbonate]. Depth - carbonate budget maintenance correlation
decreases as $\theta_{0}$ increases. This is likely due to an increasing number of reefs
being unable to maintain coral cover above threshold levels as the threshold increases, with
values becoming increasingly right-skewed. The base-10 log of total connectivity strength
was found to have no correlation with the number of years a reef maintains cover above the
required thresholds [@sfig-access-esm1-5-conn-carbonate]. Log connectivity strength -
carbonate budget maintenance correlation slightly decreases as the required carbonate budget
threshold increases.

@@

NorESM2-MM has the lowest Equilibrium Climate Sensitivity (2.5°C) and reefs exhibit
continuing recovery and decline dynamics up to 2075 under this possible climate future.
Recovery phases are predominantly seen in the ‘high’ cover cluster. The ‘medium’ cluster,
and to a lesser degree the ‘low’ cluster, also exhibit periods of recovery and increasing
proportions of coral cover up to 2045. Under NorESM2-MM the Far Northern management area has
lower CID distances between low and high cover clusters than in other management areas
[@tbl-cluster-cid], this is reflected in the low degree of cover maintenance
[@fig-gcm-cover-trajectories]. ACCESS-ESM1-5 has an intermediate ECS value of 3.87°C which
results in intermediate levels of coral cover maintenance (between those of lower and higher
ECS GCMs) [@fig-gcm-cover-trajectories], reflected in intermediate CID distance values
within most management areas [@tbl-cluster-cid]. In Townsville-Whitsunday and Far Northern
management areas, reefs in the high cluster are able to maintain coral cover levels around
20% to 2050 under ACCESS-ESM1-5. Under ACCESS-CM2 (4.66°C), EC-Earth3-Veg (4.30°C) and
GFDL-CM4 (4.1°C) GCMs cover trajectories are largely declining in all management areas,
reflected in lower CID distance values than other GCMs [@tbl-cluster-cid].

@@

Under all GCMs many reefs are projected to suffer substantial declines in coral cover
relative to 2025 levels, with an average of 38% of reefs (924 reefs) clustered into the
lowest cover level cluster at the management area scale across GCMs. NorESM2-MM and
ACCESS-ESM1-5 GCMs convey the most coral cover maintenance into the future. Differences
between low, medium and high reef clusters in cover timeseries are most notable under these
two GCMs [@tbl-cluster-cid; @fig-gcm-cover-trajectories]. Maintenance of coral cover into
the future is linked to the Equilibrium Climate Sensitivity of the GCM used as levels of
future recovery and timespan of cover maintenance increase as GCM ECS increases
[@fig-gcm-cover-trajectories]. Under high-ECS GCMs (GFDL-CM4, EC-Earth3-Veg and ACCESS-CM2)
large declines and lower rates of recovery in high cover clusters are common compared to
lower-ECS GCMs.

@@

In this investigation we model five functional groups common to the GBR, tabular Acropora,
corymbose Acropora, corymbose non-Acropora, small massives and large massives. We do not
model arborscent Acropora forms due to their distinct life history from other morphological
groups. CoralBlox uses user-defined size classes that influence size-dependent growth and
mortality rates, through which the coral colonies progress as they grow in diameter. Within
ADRIA, we define the distributions of these parameters, and the size classes based on
ecological monitoring data from the Ecological Intelligence for Reef Restoration Program
(EcoRRAP) ([@australianinstituteofmarinescience2025; @ribeirodealmeida2024]).
